\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

This project presents the development of a MATLAB simulation tool for the two-dimensional Gray--Scott reaction--diffusion model. The final program combines a shared numerical core with both non-GUI and GUI-based workflows, and includes multiple diffusion implementations, configurable boundary conditions, plotting and export utilities, reproducible verification tests, and benchmark scripts. The software was designed not only to produce Gray--Scott patterns, but also to provide a structured framework for testing solver correctness, comparing implementation variants, and supporting interactive exploration through an App Designer interface.

The numerical verification showed that the implementation reproduces the expected behavior of the underlying discretization methods. In particular, the manufactured-solution test confirmed second-order spatial accuracy of the discrete Laplacian, while the time-step study showed the expected first-order behavior of the explicit Euler scheme. Additional tests verified correct boundary-condition treatment, machine-precision agreement between matrix-based and stencil-based diffusion realizations, and consistency between GUI and non-GUI execution for the documented reference case. A Pearson-type pattern sanity check further supported the physical plausibility of the computed results.

A central practical objective of the project was to determine the most suitable solver configuration for the final implementation. The benchmark study showed that the dense formulation is mainly useful as a small-scale reference, whereas the sparse-matrix and stencil-based solvers are the practically relevant options. Among them, the stencil implementation showed the more favorable memory usage and scaling behavior over the relevant grid range, while remaining competitive in runtime. For this reason, the stencil solver was selected as the default solver in the final program. Overall, the project achieved its goal by delivering a verified, benchmarked, and usable Gray--Scott simulation tool with a modular structure suitable for further extensions.