\chapter{Introduction}

\section{Motivation}

Although the Gray--Scott model is compact in mathematical form, its numerical simulation becomes computationally demanding when spatial and temporal resolutions are increased in order to resolve pattern formation accurately. In particular, explicit time integration imposes a restrictive stability condition that links the admissible time step size to the spatial mesh width. As the grid is refined, the number of required time steps rises rapidly, which makes the model a useful test problem not only for reaction--diffusion dynamics, but also for the practical assessment of numerical efficiency \refnote{Gray--Scott references}.

A further motivation of this work is that the diffusion term can be implemented in more than one numerically equivalent way. On a structured Cartesian grid, the discrete Laplacian has a highly regular stencil structure. It can be realized explicitly through a sparse matrix construction, but it can also be applied in stencil form without assembling the full operator. This creates a natural computational question: which implementation is the more suitable practical solver for the Gray--Scott problem in the present MATLAB framework?

\section{Objective}

The objective of this project is to implement a MATLAB-based simulation tool for the two-dimensional Gray--Scott model and to determine, through systematic verification and benchmarking, which solver approach is the best practical choice for the underlying partial differential equation. The work therefore combines model implementation, numerical testing, and performance comparison in a single software framework.

A central requirement is that solver evaluation must be based on correctness first and efficiency second. For this reason, the project does not rely only on visually plausible solutions or isolated timing measurements. Instead, the implemented methods are verified for consistency and correctness before they are compared with respect to runtime and memory usage. The final aim is thus to provide a dependable simulation tool whose recommended default solver is justified by evidence rather than by assumption.